---
title: "Data Wrangling with R"
subtitle: "IBM 320 Section 02"
author: "Savannah Ventic"
date: "2026-03-04"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true

execute:
  freeze: auto
---

**Note:** This codebook is based on the content of the lecture notes published on: [Data Wrangling with Tidyverse](https://jmjung.quarto.pub/m05-1-principles-data-wrangling-with-tidyverse-in-r)

# Overview

## Learning Overview:

By the end of this module, you will be able to:

-   Describe the concept of Data Wrangling.
-   Describe how Tibbles are different from data frames
-   Explain how to convert wide or long data to “Tidy” data
-   Explain how to merge relational data sets using join functions. (Next module)
-   Explain how to use grouped mutates and filter together.
-   Be familiar with major dplyr functions for transforming data.
-   Create a new variable with mutate() and case_when().
-   Use the pipe operator to shape the data to prepare for analysis and visualization

::: {callout-tip}
To use mouse multi-cursor editing feature, do the following steps: 1. Select the text you want to edit. 2. Press `shift + alt` and click on the text you want to edit. 3. You can now edit all the selected text simultaneously.
:::

# Introduction to Data Wrangling

## Loading Packages

**Note:** To add a 'code chunk', you can use the keyboard shortcut but it depends which mode you are in.

In visual mode use: `Ctrl + Alt + I` (Windows/Linux) or `Cmd + Option + I` (Mac) in RStudio.

In source mode use: 'Ctrl + Shift + I'

```{r}
#install.packages("tidyverse")
#install.packages ("nycflights13")
library(tidyverse)
library (nycflights13)


```

## Tibbles v.s. Data Frames

```{r}
library(tibble)
library(dplyr)

class(mtcars)

cars.tib <- mtcars |>
  rownames_to_column(var = "model") |>
  as_tibble()

cars.tib
typeof(cars.tib)
```

## Data Wrangling with dplyr

```{r}
data()
```

```{r}
flights_df <- flights

flights_df |>
  count(year, month) |>
  arrange(desc(n)) |>
  mutate(month = factor(month)) |>
  ggplot(aes(x = month, y = n, fill = month)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    x = "Month",
    y = "Number of flights",
    title = "Number of flights by month in New York"
  ) +
  theme_bw()
    
```

